Hunting experience and prey variability jointly shape individual foraging specialisation in a predator-prey videogame

### 2022-10-12 manuscript/Chapter2-manuscript.Rmd
update MS and refs

*DBD* is a survival horror asymmetric multiplayer online game developed by Behavior Interactive Inc. In this game, players get to chose if they wish to play has a predator or a prey. The objective of the predator player is to search, stalk, and hunt the four prey players across a virtual environment. The objective of the four prey players is to forage for resources while avoiding the predator. The resources take the the form of power generators that once activated, will enable the prey to escape through one of two exit doors. A match ends when the predator kills all the prey available (i.e. that haven't escaped), or when the last remaining prey escapes the virtual environment. Each player, predator or prey, can chose an avatar with abilities that encourage specific play styles (e.g. bolder vs cautious prey, or ambush vs roaming predator). During our study period, the game offered 23 playable predator avatars. The virtual environment where matches take place is composed of fixed and procedurally generated habitat components, such as vegetation, mazes, and buildings. Some of these environments are larger than others, with varying structural complexity. However, we have previously shown that predators display only minimal changes in behavior and hunting success across the environments, probably due to a game feature enabling predators to have visual cues on the generators [@FraserFranco.etal2022]. There were 35 virtual game environments available for play during our study period.

Individual predators in *DBD* display variation in the hunting strategies that they use. Some specialize at hunting at high speeds and covering space in the environment, while others prefer to stalk and ambush their prey [@FraserFranco.etal2022]. They also display flexibility in the use of these tactics by switching between them across matches. The expression of these tactics and their success is highly dependent on the behavior of the prey group. The prey need to forage for resources while maintaining a high level of attention on the predator cues to avoid being detected and chased. Some prey contribute to the group's success by healing or helping others escape the predator, while other self-centered individuals play alone and attempt to escape by themselves [@Cere.etal2021]. Predators must be wary of how the prey behave to chose the right way to capture them. Thus, *DBD* simulates a highly dynamic system where both predators and prey must constantly adapt to each other to be successful.

@@

We were provided raw data by the videogame company that spanned a period of 6 months of gameplay recorded for every player. The first recorded match was played on `r toString(format(min(dat$date)))`, and the last one on `r toString(format(max(dat$date)))`. We applied cleaning and filtering procedures on the raw data to produce a synthetic dataset for our analyses. We only analyzed matches where players did not know each other (i.e. "Online" mode). We filtered any matches where players were inactive, such as when speed values were equal to, or very close to, zero. Moreover, we used our knowledge of the game to remove any matches where players were potentially hacking, or not playing the game has it is supposed to be played. These matches were not considered as player experience.

@@

All our statistical analyses were executed on Cedar (https://docs.alliancecan.ca/wiki/Cedar), a computer cluster maintained by the Digital Research Alliance of Canada. The operating system for Cedar is CentOS Linux 7. The models were fitted in R (version 4.1.2) using Markov chain Monte Carlo sampling with the package "brms" version 2.16.3 [@Burkner2017], and "cmdstanr" version 0.4.0 [@Gabry2021] as the backend for estimation (cmdstan installation version 2.28.2).

@@

We tested whether predators varied in the development of their expertise by computing 3 Bayesian generalized additive mixed models (GAMM) with thin plate regression splines that estimated the relationship between hunting success (i.e. number of prey captured) and the predators' cumulative experience (i.e. number of matches played before the current match). We parametrized the models following the method by @Pedersen.etal2019. The first model was the simplest, with a common global smoothing function for all observations and random intercepts for the predator ID. In this model, we assume that individuals have the same development of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying group-level smoothers for the predator ID. In this model, we assume that individuals share a similar relationship between success and experience, but that this relationship can vary (e.g. individual 1 has a steeper curve than individual 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the group-level smoothers for the predators, but removed the global smoother. This allows each individual to have a unique relationship between success and experience without penalization by the global smoother. Thus, this model assumes that predators do not share a common relationship between success and experience. For all three models, we controlled for the match duration.

Because a maximum of four prey can be captured in the game, we used the beta-binomial family distribution to estimate the probability of capturing all prey while controlling for overdispersion. The model parameters were estimated with a logit link function, while the dispersion parameter ($\phi$) was estimated with an identity link. The probability of capturing the four prey $P_{i}$ is drawn from a Beta distribution ($P_{i} \sim Beta(a_{i}, b_{i})$) with shape ($a_{i} = \frac{P_{i}}{\phi}$) and scale ($b_{i} = \frac{1 - P_{i}}{\phi}$) parameters. We employed the default number of basis functions (K) in "brms" for the model to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id_{0,i}$) followed a Gaussian distribution with estimated variance ($id_{0,i} \sim N(0, V_{id})$). We parametrized all models to run 4 MCMC chains with 1000 posterior samples. We ran 2500 iterations for the first model with a thinning set to 8, and 1500 iterations with a thinning set to 4 for the second and third models. The first 500 iterations were used as warmups. We provide a description on how we specified the priors for each model in the supporting information. After fitting the 3 models, we proceeded to select the one with the best predictive accuracy using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@Vehtari.etal2017].

@@

### Foraging tactics at each level of experience

We proceeded to test whether foraging behavior changes with experience, and compare the success of generalists vs specialists, using a multivariate double hierarchical linear model (MDHLM) [@@leeDoubleHierarchicalGeneralized2006;@cleasbyQuantifyingPredictabilityBehaviour2015;@westneatBiologyHiddenResidual2014;@Mitchell.etal2016a]

>>>>

*DBD* is a survival horror asymmetric multiplayer online game developed by Behavior Interactive Inc. In this game, players get to chose if they wish to play has a predator or a prey. The objective of the predator player is to hunt and capture the four prey players across a virtual environment. The objective of the four prey players is to forage for resources while avoiding the predator. The resources are in the form of power generators that once activated, will enable the prey to escape through one of two exit doors. The composition of the predator and prey group for a match is determined by a skill-based matchmaking algorithm. A match ends when the predator kills all the prey available (i.e. that haven't escaped), or when the last remaining prey escapes the virtual environment. Each player, predator or prey, can chose an avatar with abilities that encourage specific play styles (e.g. bolder vs cautious prey, or ambush vs roaming predator). During our study period, the game offered 23 playable predator avatars. The virtual environment where matches take place is composed of fixed and procedurally generated habitat components, such as vegetation, mazes, and buildings. Some of these environments are larger than others, with varying structural complexity. However, we have previously shown that predators display only minimal changes in behavior and hunting success across the environments, probably due to a game feature enabling predators to have visual cues on the generators [@FraserFranco.etal2022]. There were 35 virtual game environments available for play during our study period.

Individual predators in *DBD* display variation in the hunting strategies that they use. Some specialize at hunting at high speeds and covering space in the environment, while others prefer to stalk and ambush their prey [@FraserFranco.etal2022]. They also display flexibility in the use of these tactics by switching between them across matches. The expression of these tactics and their success is also shaped by the behavior of the prey group. The prey need to forage for resources while maintaining a high level of attention on the predator cues to avoid being detected and chased. Some prey contribute to the group's success by healing or helping others escape the predator, while other self-centered individuals play alone and attempt to escape by themselves [@Cere.etal2021]. Predators must be wary of how the prey behave to chose the right way to capture them. Thus, *DBD* simulates a highly dynamic system where both predators and prey must constantly adapt to each other to be successful.

@@

We were provided raw data by the videogame company that spanned a period of 6 months of gameplay recorded for every player. The first recorded match was played on `r toString(format(min(dat$date)))`, and the last one on `r toString(format(max(dat$date)))`. We applied cleaning and filtering procedures on the raw data to produce a synthetic dataset for our analyses. We only analyzed matches where players did not know each other (i.e. "Online" mode). We filtered any matches where players were inactive, such as when speed values were equal to, or very close to, zero. Moreover, we used our knowledge of the game to remove any matches where players were potentially hacking, or not playing the game has it is supposed to be played. These matches were not considered as player experience. We then proceeded to partition the player population by total experience, and randomly sampled players to have a uniform representation of total gameplay experience. To optimize the representation of gameplay experience at advanced levels, we set the maximum amount of matches at 500 for these players. For instance, there could be a large difference in gameplay between 300-500 and 500-1000 matches. Players that played above 500 matches represented only ~2% of the population.

@@

All our statistical analyses were executed on Cedar (https://docs.alliancecan.ca/wiki/Cedar), a computer cluster maintained by the Digital Research Alliance of Canada. The operating system for Cedar is CentOS Linux 7. The models were fitted in R (version 4.1.2) using Markov chain Monte Carlo sampling with the package "brms" version 2.16.3 [@Burkner2017], and "cmdstanr" version 0.4.0 [@Gabry.Cesnovar2021] as the backend for estimation (cmdstan installation version 2.28.2).

@@

We tested whether predators varied in the development of their expertise by computing 3 Bayesian generalized additive mixed models (GAMM) with thin plate regression splines estimating the relationship between hunting success (i.e. number of prey captured) and the predators' cumulative experience (i.e. number of matches played before the current match). We parametrized the models following the method by @Pedersen.etal2019. The first model was the simplest, with a common global smoothing function for all observations and random intercepts for the predator ID. In this model, we assume that individuals have the same development of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying group-level smoothers for the predator ID. In this model, we assume that individuals share a similar relationship between success and experience, but that this relationship can vary (e.g. individual 1 has a steeper curve than individual 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the group-level smoothers for the predators, but removed the global smoother. This allows each individual to have a unique relationship between success and experience without penalization by the global smoother. Thus, this model assumes that predators do not share a common relationship between success and experience. For all three models, we controlled for the match duration.

Because a maximum of four prey can be captured in the game, we used the beta-binomial family distribution to estimate the probability of capturing all prey while controlling for overdispersion. The model parameters were estimated with a logit link function, while the dispersion parameter ($\phi$) was estimated with an identity link for all models. The probability of capturing the four prey $P_{i}$ is drawn from a Beta distribution ($P_{i} \sim Beta(a_{i}, b_{i})$) with shape ($a_{i} = \frac{P_{i}}{\phi}$) and scale ($b_{i} = \frac{1 - P_{i}}{\phi}$) parameters. We employed the default number of basis functions (K) in "brms" for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id_{0,i}$) followed a Gaussian distribution with estimated variance ($id_{0,i} \sim N(0, V_{id})$). We parametrized all models to run 4 MCMC chains with 1000 posterior samples. We ran 2500 iterations for the first model with a thinning set to 8, and 1500 iterations with a thinning set to 4 for the second and third models. The first 500 iterations were used as warmups. We provide a description on how we specified the priors for each model in the supporting information. After fitting the 3 models, we proceeded to select the one with the best predictive accuracy using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@Vehtari.etal2017].

@@

### Foraging behavior and success, prey behavior, and their relationships, at each level of experience

We proceeded to test whether foraging behavior changes with experience, and compare the success of generalists vs specialists, using a multivariate double hierarchical linear model (MDHLM) [@leeDoubleHierarchicalGeneralized2006;@cleasbyQuantifyingPredictabilityBehaviour2015;@westneatBiologyHiddenResidual2014;@Mitchell.etal2016a;@ODea.etal2022].

expliquer comment j'ai partitionné l'expérience basée sur les résultats du GAMM
expliquer que j'ai calculé le coefficient de variation pour comparer le comportement
